Tracing the Meta-Level: PyPy's JIT Compiler

### 2009-03-31 talk/icooolps2009/paper.tex
more comments

@@

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The hot loop of the language
interpreter is the bytecode dispatch loop. Usually it is also the only hot loop.
Tracing one iteration of this loop means that the execution of one bytecode was
seen. This means that the resulting machine code will correspond to a loop, that
assumes that this particular bytecode will be executed many times in a row,
which is clearly very unlikely.

@@

Let's look at an example. Figure \ref{fig:tlr-basic} shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator. The
\texttt{bytecode} argument is a string of bytes and all register and the
accumulator are integers. A simple program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} bytecode, the
trace would look as follows:
\input{code/normal-tracing.txt}

To improve this situation, the tracing JIT could trace the execution of several
bytecodes, thus effectively unrolling the bytecode dispatch loop. Ideally, the
bytecode loop should be unrolled exactly so much, that the unrolled version
corresponds to a loop on the level of the user program. A loop in the user
program occurs when the program counter of the language interpreter has the
same value many times. This program counter is typically one or several
variables in the language interpreter, for example the bytecode object of the
currently executed function of the user program and the position of the current
bytecode within that.

@@

The critical problem of tracing the execution of just one bytecode has been
solved, the loop corresponds exactly to the loop in the square function.
However, the resulting trace is a bit too long. Most of its operations are not
actually doing any computation that is part of the square function. Instead,
they manipulate the data structures of the language interpreter. While this is
to be expected, given that the tracing interpreter looks at the execution of the
language interpreter, it would still be nicer if some of these operations could
be removed.

@@

\begin{figure}
\input{code/full.txt}
\caption{Trace when executing the Square function of Figure \ref{fig:square},
with the corresponding bytecodes as comments. The constant-folding of operations
on green variables is enabled.}
\label{fig:trace-full}
\end{figure}

>>>>

\normalem

@@

\anto{I find this para a bit confusing, but I don't really know how to improve
  it. I try a rewriting, feel free to revert}

\sout{A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The hot loop of the language
interpreter is the bytecode dispatch loop. Usually it is also the only hot loop.
Tracing one iteration of this loop means that the execution of one opcode was
seen. This means that the resulting machine code will correspond to a loop, that
assumes that this particular opcode will be executed many times in a row,
which is clearly very unlikely.}

\anto{Similarly, we need to distinguish two kind of loops: \emph{interpreter
    loops} are loops \textbf{inside} the language interpreter (e.g., the
  bytecode dispatch loop). \emph{User loops} are loops in the user program.}

\anto{To improve the performances of the user programs, we need to find and
  compile traces that correspond to user loops.  The strategy described in
  section \ref{sect:tracing} does not work well, as it only detects
  interpreter loops, and in particular the bytecode dispatch one.}

@@

Let's look at an example. Figure \ref{fig:tlr-basic} shows the code of a very
simple bytecode interpreter with 256 registers and an accumulator. The
\texttt{bytecode} argument is a string of bytes and all register and the
accumulator are integers. A simple program for this interpreter that computes
the square of the accumulator is shown in Figure \ref{fig:square}. If the
tracing interpreter traces the execution of the \texttt{DECR\_A} opcode (whose integer value is 7), the
trace would look as follows:
\input{code/normal-tracing.txt}

\anto{ Because of the guard on \texttt{opcode0}, the code compiled from this
  trace will be useful only when executing a \texttt{DECR\_A}.  For all the
  other operations the guard will fail, making the whole system probably
  slower than without JIT.}

% YYY: (anto) I reviewd until here

To improve this situation, the tracing JIT could trace the execution of several
opcodes, thus effectively unrolling the bytecode dispatch loop. Ideally, the
bytecode dispatch loop should be unrolled exactly so much, that the unrolled version
corresponds to a loop on the level of the user program. A loop in the user
program occurs when the program counter of the language interpreter has the
same value many times. This program counter is typically one or several
variables in the language interpreter, for example the bytecode object of the
currently executed function of the user program and the position of the current
bytecode within that.

@@

The critical problem of tracing the execution of just one opcode has been
solved, the loop corresponds exactly to the loop in the square function.
However, the resulting trace is a bit too long. Most of its operations are not
actually doing any computation that is part of the square function. Instead,
they manipulate the data structures of the language interpreter. While this is
to be expected, given that the tracing interpreter looks at the execution of the
language interpreter, it would still be nicer if some of these operations could
be removed.

@@

\begin{figure}
\input{code/full.txt}
\caption{Trace when executing the Square function of Figure \ref{fig:square},
with the corresponding opcodes as comments. The constant-folding of operations
on green variables is enabled.}
\label{fig:trace-full}
\end{figure}
